Translation of B Implementations to the LLVM: Control and Modularity

### 2013-09-25 translation.tex
Remodeled the definition of code generation for imports.

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

        \B{impo} & : & seq \B{Module} \\
        \B{op} & : & seq \B{Oper} 
        \multicolumn{3}{|c|}{Modules: \B{Module}} \\
        \B{mach} & : & \B{MachRef} \\
        \multicolumn{3}{|c|}{Machine: \B{MachRef}} \\
        \B{spec} & : & \B{MachInterface} \\
        \multicolumn{3}{|c|}{Machine interface: \B{MachInterface}} \\

@@

        \B{op} & : & seq \B{Oper} 

@@

In Atelier-B, software development is organized in libraries of modules. A
module has a specification, called a machine. A machine may be:
\begin{description}
\item[basic] In that case its implementation is carried out outside the realm of
  the B method. A possible application is a set of related I/O routines. Even
  though no LLVM code needs to be generated to implement the basic machines, the
  code generator needs still to produce LLVM declarations of the entities
  (e.g., constants and operations) that are supplied by those machines.
\item[developed] In that case, this module also contains a series of
  refinements, eventually reaching an implementable version, from which LLVM
  code may be generated.
\end{description}
Amongst the several constructs of the B notation supporting modular design, the
\emph{import} relation is related to separate compilation. At the implementation
level, one module may import several instances of a machine to construct its
internal data structures. Such imported machines may be either basic, or
developed. In the latter case, the implementation of the imported machines may
in turn import other machine instances and there is no pre-established limit to
such chain of imports. Hence, the import relation between module instances forms
a tree, where the root is an implementation and the descendants of a node are
the instances imported from the module in that node.

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{b-project.pdf}
\caption{Structure of the imports relation in a B project}
\label{fig:structure}
\end{center}
\end{figure}
Figure~\ref{fig:structure}
illustrates this with the structure of a B project where an implementation
\B{I0} of a specification \B{M0} imports one unnamed instance of a basic machine
\B{M2} and two instances named \B{A} and \B{B} of a developed machine \B{M1}. In
turn its implementation \B{I1} itself imports two instances \B{C} and \B{D} of a
developed machine \B{M3}, implemented as \B{I3} (which has no imports itself).
The generation of LLVM code for the project shall allocate memory for one
instance of \B{I0}, two instances of \B{I2} and four instances of \B{I3}.

At that point, let us remark that, when generating code for a given module, we
need to consider the following two cases:
\begin{enumerate}
\item The code generation aims at producing a component, that will further be
  instantiated by importations. We call this the \emph{COMP} mode for code
  generation. This case is further split into two sub-cases:
  \begin{enumerate}
  \item The module is basic, and only declarations need to be emitted.
  \item The module is developed, and declarations and definitions are produced.
  \end{enumerate}
\item The generated code corresponds to a full project. Given the root module,
  the code generator must identify all the transitively imported module instances. We call this the \emph{PROJ} mode for code generation.
\end{enumerate}
One key issue in the code generation is how memory shall be allocated to store
the state of the modules. To satisfy our first initial requirement (no dynamic
memory allocation), this allocation shall be static, that is to say the
generator must include the code that creates the module instances when the
project starts execution. Note that this cannot be done in \emph{COMP} mode as,
at that point, we cannot assume any knowledge on the number of instances the
module being compiled we there be at run time. Consequently, the responsibility
for the static allocation of instances is assigned to the \emph{PROJ} mode,
which shall also perform the proper bindings between the actual and formal
instances in the generated code.

To encode the data space of a module, the translation produces a structure type,
named \llvm{\%M\$state\$}, such that the elements of the structure represent the
variables and the instances of imported modules. To satisfy our second initial
requirement (separate compilation), the instances cannot be part of the
structure itself, instead references to those instances (encoded as pointers)
are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
To encode behavior of a module, each operation \B{op} is translated into the
definition of a function named \llvm{@M\$op}The parameter list of the functions
thus produced contains one item for each input and output of the corresponding
operation. There is also one parameter of type \llvm{\%M\$ref\$}, to pass the
address of the structure encoding instance associated with that operation. Also
the translation produces a special function \llvm{@M\$init\$} responsible for
the initialization. The parameters of this function are the addresses of the
instances found in the import tree of the module (including the module itself).

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{description}
\item[typedef:] Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module. In case the data space of a module is empty,
  this section is empty and the type is not defined:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data}, if
  this has been defined. 
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of LLVM
  functions: one implementing the initialisation of the module, and one for each
  operation of the module. If the data space of the module is empty, no function
  definition corresponding to the initialisation is emitted
  \begin{tabbing}
    foo \= \kill
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item[implementation:] Defines the functions declared in the interface. This
  section is produced only for the developed modules.
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
\end{description}
Assuming these three sections are available, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \textit{generate a type declaration for the state space of} \B{Q} \\
    \> \llvm{\%Q\$state\$ = type opaque} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

@@

The detailed presentation of the translation of the different elements of a B
implementation is divided into:
\begin{description}
\item[Modularity] constructs to combine B components are described in
  sec.~\ref{sec:module}.
\item[Data] translation is briefly described sec.~\ref{sec:data}, including
  the encoding of B data types.
\item[Expressions] and conditions are codified as LLVM instruction blocks, as
  described in section~\ref{sec:expr}.
\item[Control] flow, instructions are translated to LLVM instructions as
  discussed in sec.~\ref{sec:control}, which also presents the encoding
  of B operations and initialization.
\end{description}

\paragraph{Notations.} Code generation is defined with a set of rules defined by
structural recursion. For each grammatical construct, a code generation function
is defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
instructions, $\trad{~}{\B{Expr}}$ for expressions and so forth. Each such
function usually produces a text corresponding to bits of LLVM source code. It
may have additional parameter and results. The symbol $\sep$ is used to clearly
separate composed results. Auxiliary functions are introduced for
conveniency. To abbreviate the presentation, we denote
$\trad{~}{\ListOf{\B{A}}}$ as the translation function for sequences of \B{A}
elements, defined as the concatenation of the application of $\trad{~}{\B{A}}$
to the elements of the sequence.

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name.  Also, we define
three auxiliary functions responsible to produce such identifiers:
\begin{align*}
\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
\end{align*}
We assume an unlimited supply of labels and local variables: we use $\lbl$ to
return a fresh LLVM label and $\variable$ for a fresh local variable name.

Finally, we need a function that, given a concrete variable, or an imported
machine, returns the position of the corresponding element in the structure
representing the state of the implementation. The position of the elements in
the structure is the same as the position of corresponding B construct in the
source code:
\begin{align*}
  \idx(\B{n}) & \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
\B{n.root.vari})
\end{align*}
where $\mathit{index}(e, s)$ is a function that returns the position of $e$ in
sequence $s$ and $\conc$ is the concatenation operator for sequences.

@@

[[[ THIS SECTION IS DEPRECATED ]]]

The translation of a B implementation to LLVM, following the template sketched
in Figure~\ref{fig:skel}, is specified by the following rule, where \B{n} is the
root of the implementation representation and $\top^?$ a boolean indicating if
\B{n} is the implementation of a top-level component or of a library machine:
\begin{align*}
  \lefteqn{\trad{\B{n}, $\top^?$}{\B{Impl}} \isdef } \\
  & \llvm{"\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{State}}}_{def}}} & \text{defines the type of imported modules} \\
  & \llvm{\state(\B{n}) = type \PH{\trad{n}{State}_{expr}}} & \text{defines the type of current module} \\
  & \llvm{\PH{\trad{\B{n.impo}, $\emptyset$}{\ListOf{\B{Bhv}}}_{decl}}} & \text{declares of imported modules operations} \\
  & \llvm{\PH{\trad{\B{n.init}}{\B{Init}}}} & \text{defines the initialization function} \\
  & \llvm{\PH{\trad{\B{n.op}}{\ListOf{\B{Oper}}}}} & \text{defines functions implementing operations}\\
  & \llvm{\PH{\trad{\B{n}, $\top^?$}{\B{Inst}}}"} & \text{creates instance (conditional)}
\end{align*}

To handle the importation of a machine, one needs to produce the definition of
the type for the state of its implementation. Since this implementation may reference
other modules, the definition is recursive (parameter $A$ is an accumulator to avoid
duplicating such definitions):
\begin{align*}
  \trad{\B{n}, $A$}{\B{State}}_{def} \isdef 
  & \IF \B{n} \not\in A \THEN \\
  & \quad \LET \B{m} = \B{n.mach} \IN \\
  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
  & \quad \quad \quad \llvm{"\PH{\trad{\B{i.impo}, $A \cup \{ \B{n} \}$}{\ListOf{\B{State}}}_{def}}} \\
  & \quad \quad \quad \llvm{\PH{\state(\B{m})} = type \PH{\trad{i}{State}_{expr}} \nl"}
\end{align*}

The following rules specify how the type representing the state of an
implementation is generated. It is a structure type, with one element per
imported machine, and one element per variable. The definition of such type is
just the enumeration of the type of its elements (the individual elements are
referenced, not by name, but by position inside the structure):
\begin{align*}
  \trad{\B{n}}{\B{State}}_{expr} \isdef & \llvm{"\{ \PH{\trad{\B{n.impo}}{\ListOf{\B{Impo}}}_{expr}},  \PH{\trad{\B{n.var}}{\ListOf{\B{Vari}}}_{expr}} \}"} \\
  \trad{\B{n}}{\B{Impo}}_{expr} \isdef & \state(\B{n.mach}) \\
  \trad{\B{n}}{\B{Vari}}_{expr} \isdef & \trad{\B{n.type}}{\B{Type}}
\end{align*}

The code generation for the functions implementing the initialization and
operations of imported machine is specified by the following rules. Not that
this definition needs not be recursive, as only the operations of a directly
imported machine may be called in an implementation. Again, parameter $A$ plays
the role of an accumulator to avoid (harmless, but useless) duplicate
declarations.
\begin{align*}
  \trad{\B{n}, $A$}{\B{Bhv}}_{decl} \isdef 
  & \IF \B{n} \not\in A \THEN \\
  & \quad \LET \B{m} = \B{n.mach} \IN \\
  & \quad \quad \LET \B{i} = \B{m.impl} \IN \\
  & \quad \quad \quad \llvm{declare void \PH{\init(\B{m})}(\PH{\state(\B{m})} *) \nl} \\
  & \quad \quad \quad \llvm{\PH{\trad{\B{m.op}}{\ListOf{\B{Oper}}}_{decl}}"} \\
  \trad{\B{n}}{\B{Oper}}_{decl} \isdef &
  \llvm{"declare void \PH{\Global(\B{n})}(\PH{\state(\B{n.root})}, \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}}, \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}})"}\\
  \trad{\B{n}}{\B{inp}}_{decl} \isdef & \trad{\B{n.type}}{\B{Type}} \\
  \trad{\B{n}}{\B{out}}_{decl} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
\end{align*}

The last rule is responsible for producing the code creating the instance when the
implementation is that of the top-level component of a design:
\begin{align*}
\lefteqn{  \trad{\B{n}, $\top^?$}{\B{Inst}} \isdef } \\
  & \IF \top^? \THEN \llvm{"\PH{\self(\B{n})} = common global \PH{\state(\B{n})} zeroinitializer \nl"} \\
  & \ELSE \llvm{""}
\end{align*}

>>>>

\newcommand{\trad}[2]{\ensuremath{\lVert \textsf{#1} \rVert^{\textit{#2}}}}
\newcommand{\nl}[0]{\ensuremath{\downarrow}}
\newcommand{\mty}[0]{\texttt{""}}
\DeclareMathOperator{\conc}{\diamond}
\DeclareMathOperator{\isdef}{\equiv}
\DeclareMathOperator{\dom}{\mbox{dom}}
\DeclareMathOperator{\lbl}{\mathcal{L}()}
\DeclareMathOperator{\variable}{\mathcal{V}()}
\newcommand{\llvm}[1]{\texttt{#1}}
\newcommand{\B}[1]{\textsf{#1}}
\newcommand{\lalt}[0]{$\langle$\xspace}
\newcommand{\ralt}[0]{$\rangle$\xspace}
\newcommand{\alt}[0]{$\mid\,$}
\newcommand{\ListOf}[1]{$\mbox{#1}^+$}
\newcommand{\nt}[1]{{\normalfont\textit{#1}}}
\newcommand{\Dict}[0]{\mathbb{D}}
\newcommand{\Text}[0]{\mathbb{T}}
\newcommand{\IF}[0]{\textbf{ if }}
\newcommand{\ELSIF}[0]{\textbf{ else if }}
\newcommand{\ELSE}[0]{\textbf{ else }}
\newcommand{\THEN}[0]{\textbf{ then }}
\newcommand{\LET}[0]{\textbf{ let }}
\newcommand{\IN}[0]{\textbf{ in }}
\newcommand{\AND}[0]{\textbf{ and }}
\newcommand{\PH}[1]{\framebox{$#1$}}
\newcommand{\sep}[0]{\otimes}
\newcommand{\intf}[0]{\ensuremath{\mathbb{I}}}
\newcommand{\Global}[0]{\ensuremath{\sf\Gamma}}
\newcommand{\local}[0]{\ensuremath{\sf\lambda}}
\newcommand{\opmap}[0]{\ensuremath{\sf\Omega}}
\newcommand{\developed}[0]{\ensuremath{\textsf{developed}}}
\newcommand{\stateless}[0]{\ensuremath{\textsf{stateless}}}
\newcommand{\importedmodules}[0]{\ensuremath{\textsf{i-mod}}}
\newcommand{\trimportedmodules}[0]{\ensuremath{\textsf{i-mod$\mathsf{\ast}$}}}
\newcommand{\importedinstances}[0]{\ensuremath{\textsf{i-inst}}}
\newcommand{\trimportedinstances}[0]{\ensuremath{\textsf{i-inst$\mathsf{\ast}$}}}
\newcommand{\idx}[0]{\ensuremath{\sf\Pi}}
\newcommand{\state}[0]{\ensuremath{\sf\Theta}}
\newcommand{\stateref}[0]{\ensuremath{\sf\Phi}}
\newcommand{\self}[0]{\ensuremath{\sf\Sigma}}
\newcommand{\init}[0]{\ensuremath{\mathsf{I}}}
\newcommand{\tradi}[2]{\ensuremath{\langle \textsf{#1} \rangle^{\textit{#2}}}}

@@

        \multicolumn{3}{|c|}{Machine: \B{Machine}} \\

@@

        \B{op} & : & seq \B{Oper} \\
        \B{impo} & : & seq \B{Import} \\
        \B{op} & : & seq \B{Oper} \\
        \multicolumn{3}{|c|}{Imports: \B{Import}} \\
        \B{mach} & : & \B{Machine} \\

@@

In Atelier-B, software development is organized in libraries of modules. A
module has a specification, called a machine. A machine may be:
\begin{description}
\item[basic] In that case its implementation is carried out outside the realm of
  the B method. Even though no LLVM code needs to be generated to implement the
  basic machines, the code generator still needs to produce LLVM declarations of
  the entities (e.g., constants and operations) that are supplied by those
  machines.
\item[developed] In that case, this module also contains a series of
  refinements, eventually reaching an implementable version, from which LLVM
  code may be generated.
\end{description}
We distinguish basic machines from developed machines in the abstract syntax by
testing the \B{impl} attribute of the \B{Machine} element: if it is $\bot$, then
the machine is basic, otherwise it is the root of the corresponding
implementation.

Amongst the several constructs of the B notation supporting modular design, the
\emph{import} relation is related to separate compilation. At the implementation
level, one module may import several instances of a machine to construct its
internal data structures. Such imported machines may be either basic, or
developed. In the latter case, the implementation of the imported machines may
in turn import other machine instances and there is no pre-established limit to
such chain of imports. Hence, the import relation between module instances forms
a tree, where the root is an implementation and the descendants of a node are
the instances imported from the module in that node.

\begin{figure}
\begin{center}
\includegraphics[width=.8\textwidth]{b-project.pdf}
\caption{Structure of the imports relation in a B project}
\label{fig:structure}
\end{center}
\end{figure}
Figure~\ref{fig:structure} illustrates this with the structure of a B project
where an implementation \B{I0} of a specification \B{M0} imports one unnamed
instance of a basic machine \B{M1} and two instances named \B{A} and \B{B} of a
developed machine \B{M2}. Its implementation \B{I2} in turn imports two
instances \B{C} and \B{D} of a developed machine \B{M3}, implemented as \B{I3}
(which has no imports itself).  The generation of LLVM code for the project
shall produce the representation for one instance of \B{I0}, two instances of
\B{I2} and four instances of \B{I3}.

At that point, let us remark that, when generating code for a given module, we
need to consider the following two cases:
\begin{enumerate}
\item The code generation aims at producing a component, that will further be
  instantiated by importations. We call this the \emph{COMP} mode for code
  generation. This case is further split into two sub-cases:
  \begin{enumerate}
  \item The module is basic, and only declarations need to be emitted.
  \item The module is developed, and declarations and definitions are produced.
  \end{enumerate}
\item The generated code corresponds to a full project. Given the root module,
  the code generator must identify all the transitively imported module instances. We call this the \emph{PROJ} mode for code generation.
\end{enumerate}
One key issue in the code generation is how memory shall be allocated to store
the state of the modules. To satisfy our first initial requirement (no dynamic
memory allocation), this allocation shall be static, that is to say the
generator must include the code that creates the module instances when the
project starts execution. Note that this cannot be done in \emph{COMP} mode as,
at that point, we cannot assume any knowledge on the number of instances the
module being compiled will there be at run time. Consequently, the
responsibility for the static allocation of instances is assigned to the
\emph{PROJ} mode, which shall also perform the proper bindings between the
actual and formal instances in the generated code.

To encode the data space of a module, the translation produces a structure type,
named \llvm{\%M\$state\$}, such that the elements of the structure represent the
variables and the instances of imported modules. To satisfy our second initial
requirement (separate compilation), the instances cannot be part of the
structure itself, instead references to those instances (encoded as pointers)
are used. Let \llvm{\%M\$ref\$} denote the type pointer to \llvm{\%M\$state\$}.
To encode behavior of a module, each operation \B{op} is translated into the
definition of a function named \llvm{@M\$op}. The parameter list of the
functions thus produced contains one item for each input and output of the
corresponding operation. There is also one parameter of type \llvm{\%M\$ref\$},
to pass the address of the structure encoding instance associated with that
operation. Also the translation produces a special function \llvm{@M\$init\$}
responsible for executing the initialization of \B{M}. The parameters of this
function are the addresses of the instances found in the import tree of the
module (including the module itself).

Our approach consists in dividing the translation into three sections, and to
combine these sections differently in \emph{PROJ\/} and \emph{COMP\/} mode. For
a given module \B{M}, the sections are the following:
\begin{description}
\item[typedef:] Defines an LLVM data type \llvm{\%M\$state\$} corresponding to
  the data space of the module:
  \begin{tabbing}
    \llvm{\%M\$state\$ = type \{ \ListOf{\nt{type}} \}}
  \end{tabbing}
  In case the data space of a module is empty, this section is empty and the
  type is not defined.
\item[interface:] Defines an LLVM type \B{M\_ref}, pointer to \B{M\_data} (if
  defined):
  \begin{tabbing}
    \llvm{\%M\$ref\$ = type \%M\$state\$*}
  \end{tabbing}
  The interface also contains declarations of LLVM
  functions: one implementing the initialisation of the module, and one for each
  operation of the module. If the data space of the module is empty, no function
  definition corresponding to the initialisation is emitted:
  \begin{tabbing}
    foo \= \kill
    \llvm{declare void @M\$op(\%M\$ref\$, \ListOf{\nt{type}})} \\
    \llvm{declare void @M\$init\$(\%M\$ref\$, \ListOf{\nt{type}})}
  \end{tabbing}
\item[implementation:] Defines the functions declared in the interface:
  \begin{tabbing}
    foo \= \kill
    \llvm{define void @M\$op(\%M\$state\$* \%self\$, \ListOf{\nt{param}}) \{} \\
    \> \llvm{\ListOf{\nt{block}}} \\
    \> \llvm{exit: ret void} \\
    \llvm{\}}
  \end{tabbing}
  This section is produced only for the developed modules.
\end{description}
Based on the definition of these sections, the \emph{COMP\/} translation
assembles them as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each imported module} \B{Q} \textit{generate} \\
    \> \llvm{\%Q\$state\$ = type opaque} \quad \textit{(declares type for} \B{Q} \textit{state space)} \\
    \> \textit{include the interface section of }\B{Q} \\
    \textit{include the typedef section of} \B{M} \\
    \textit{include the implementation section of} \B{M}
  \end{tabbing}
\end{center}
For a module \B{M}, the \emph{PROJ\/} translation is organized as follows:
\begin{center}
  \begin{tabbing}
    foo \= foo \= \kill
    \textit{for each transitively imported component} \B{Q} \\
    \> \textit{include the typedef section of \B{Q}} \\
    \textit{for each instance} \B{Q} \textit{imported transitively through} \B{path} \\
    \> \textit{declare a variable of type} \llvm{\%Q\$state\$}: \\
    \> \llvm{@Q$\lbrack$path$\rbrack$ = common global \%M\$state\$ zeroinitializer} \\

@@

The detailed presentation of the translation of the different elements of a B
implementation is divided into:
\begin{description}
\item[Notation] and some auxiliary definitions are presented in sec.~\ref{sec:notation}.
\item[Modularity] constructs to combine B components are described in
  sec.~\ref{sec:module}.
\item[Data] translation is briefly described sec.~\ref{sec:data}, including
  the encoding of B data types.
\item[Expressions] and conditions are codified as LLVM instruction blocks, as
  described in section~\ref{sec:expr}.
\item[Control] flow, instructions are translated to LLVM instructions as
  discussed in sec.~\ref{sec:control}, which also presents the encoding
  of B operations and initialization.
\end{description}

\section{Notations} 
\label{sec:notation}

Code generation is defined with a set of rules defined by
structural recursion. For each grammatical construct, a code generation function
is defined: $\trad{~}{\B{Oper}}$ for operations, $\trad{~}{\B{Inst}}$ for
instructions, $\trad{~}{\B{Expr}}$ for expressions and so forth. Each such
function usually produces a text corresponding to bits of LLVM source code. It
may have additional parameter and results. The symbol $\sep$ is used to clearly
separate composed results. Auxiliary functions are introduced for
conveniency. To abbreviate the presentation, we denote
$\trad{~}{\ListOf{\B{A}}}$ as the translation function for sequences of \B{A}
elements, defined as the concatenation of the application of $\trad{~}{\B{A}}$
to the elements of the sequence.

In those rules, $\PH{t}$ indicates a text place-holder where the value of $t$
shall be inserted and $\nl$ denotes a new line. For instance, we define
$\local(\B{n}) \isdef \llvm{"\%\PH{\B{n.id}}"}$ as a function that given an
element \B{n} of the abstract syntax tree returns the corresponding LLVM local
name, obtained by prepending character \% to its identifier; and $\Global(\B{n})
\isdef \llvm{"@\PH{\B{n.root.id}}\$\PH{\B{n.id}}"}$ yields the corresponding
LLVM global name, composed by special characters (to avoid name conflicts), the
name of the component where it is defined and its own name.  Also, we define
additional auxiliary functions responsible for producing such identifiers:
\begin{align*}
\state(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$state\$"} & \text{ (the type for the state)} \\
\stateref(\B{n}) & \isdef \llvm{"\%\PH{\B{n.id}}\$ref\$"} & \text{ (the type pointer to state)} \\
\init(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$init\$"} & \text{ (the initialization function)} \\
\self(\B{n}) & \isdef \llvm{"@\PH{\B{n.id}}\$self\$"} & \text{ (the instance in a top-level module)}
\end{align*}
We assume an unlimited supply of labels and local variables: we use $\lbl$ to
return a fresh LLVM label and $\variable$ for a fresh local variable name. Also, let us define:
\begin{itemize}
\item $\developed(\B{n}) \isdef \B{n.impl} \neq \bot$ to test if a machine is
  developed.
\item $\stateless(\B{n}) \isdef \B{n.vars} =  \bot \land \forall i \in \B{n.impo}, \stateless(i)$ to test if a machine is stateless.
\item $\importedmodules(\B{n}) \isdef \{ \B{i.mach} \mid \B{i} \in \B{n.impo}
  \}$ to get the set of machines directly imported from implementation $\B{n}$.
\item $\importedinstances(\B{n}) \isdef \B{n.impo}$ to get the set of machines
  instances directly imported from implementation $\B{n}$.
\item $\trimportedmodules(\B{n}) \isdef \ldots$ to get the set of machines
transitively imported from implementation $\B{n}$.
\item $\trimportedinstances(\B{n}) \isdef \ldots$ to get the set of machines
instances transitively imported from implementation $\B{n}$.
\item $\idx(\B{n}) \isdef \mathit{index}(\B{n}, \B{n.root.impo} \conc
  \B{n.root.vari})$ to get, from a given concrete variable $\B{n}$, the position
  of the corresponding element in the structure representing the state of the
  implementation. Here where $\mathit{index}(e, s)$ is a function that returns
  the position of $e$ in sequence $s$ and $\conc$ is the concatenation operator
  for sequences.
\end{itemize}

@@

\subsection{The typedef section}

This section corresponds to the LLVM type definition for the state space of a
given module. If the module is developed, then the state space is that of the
corresponding implementation, otherwise the module is basic. If the basic
machine is stateless, the state space is empty; otherwise, it is represented by
an aggregate type, the elements of which correspond to the variables of the
machine:
\begin{align*}
  \trad{\B{n}}{\B{Mach}}_{\text{typedef}} \isdef 
  & \IF \developed(\B{n}) \THEN \trad{\B{n.impl}, \B{n}}{\B{Impl}}_{\text{typedef}} \\
  & \ELSE \IF \stateless(\B{n}) \THEN \llvm{""} \\
  & \ELSE \llvm{"\PH{\state(\B{n})} = type \{ \PH{\trad{\B{n.vars}}{\ListOf{\B{Vari}}}_{\text{typedef}}} \} "}
\end{align*}

For the case of a developed machine \B{m} with implementation \B{n}, if \B{n} is
stateless, no type definition is emitted; otherwise, it is an aggregate type
composed of one element for each import and variable:
\begin{align*}
  \trad{\B{n}, \B{m}}{\B{Impl}}_{\text{typedef}} \isdef 
  & \IF \stateless(\B{n}) \THEN \llvm{""} \\
  & \ELSE \llvm{"\PH{\state(\B{m})} = type \{ \PH{\trad{\B{n.impo}}{\ListOf{\B{Impo}}}_{\text{typedef}}}, \PH{\trad{\B{n.var}}{\ListOf{\B{Vari}}}_{\text{typedef}}} \} "}
\end{align*}

In these aggregate types, the elements corresponding to an import \B{n} is
defined in the following rule. If the imported machine is stateless, there is no
element, otherwise it is the name of the type pointer to the state of the
machine:
\begin{align*}
  \trad{\B{n}}{\B{Impo}}_{\text{typedef}} \isdef 
  & \llvm{""} \triangleleft \stateless(\B{n.mach}) \triangleright \stateref(\B{n.mach})
\end{align*}

The elements corresponding to concrete variables are the LLVM data types
corresponding to the B types of those variables (defined in
sec.~\ref{sec:data}):
\begin{align*}
  \trad{\B{n}}{\B{Impl}}_{\text{typedef}} \isdef 
  & \trad{\B{n.type}}{\B{Type}}
\end{align*}

\subsection{The interface section}

The interface section of a module \B{n} defines a type $\stateref(\B{n})$,
pointer to $\state(\B{n})$, the type to represent the data space of \B{n}, and
declares functions for executing 1) executing the initialization of an instance
of \B{n} and 2) each operation in \B{n}:
\begin{align*}
  \trad{\B{n}}{\B{Mach}}_{\text{interface}} \isdef 
  & \llvm{"\PH{\stateref(\B{n})} = type \PH{\state(\B{n})}*\nl} \\
  & \llvm{\PH{\trad{\B{n}}{\B{init}}_{\text{interface}}}}\\
  & \llvm{\PH{\trad{\B{n.op}}{\ListOf{\B{Oper}}}_{\text{interface}}}"}
\end{align*}

The function responsible for initializing an instance of \B{n} takes as
parameter a reference to such instance, as well as a reference for each
transitively imported stateful instance and is as follows:
\begin{align*}
  \trad{\B{n}}{\B{init}}_{\text{interface}} \isdef 
  & \llvm{"declare void \PH{\init(\B{n})} ( \PH{\stateref(\B{n})}, 
    \PH{\trad{\trimportedinstances(\B{n})}{\ListOf{\B{inst}}}_{\text{decl}}} )\nl"} \\
  \trad{\B{n}}{\B{inst}}_{\text{decl}} \isdef 
  & \bot \triangleleft \stateless(\B{n.mach}) \triangleright \stateref(\B{n.mach}) 
\end{align*}

The function responsible for executing an operation \B{n} has as parameter a
reference to the instance of the corresponding module, if it is stateful, and
one parameter for each input and output of the corresponding operation:
\begin{align*}
  \trad{\B{n}}{\B{oper}}_{\text{interface}} \isdef 
  & 
\LET t = \bot \triangleleft \stateless(\B{n.root}) \triangleright \stateref(\B{n.root}) \IN \\
& \llvm{"declare void \PH{\Global(\B{n})} 
( \PH{t}, 
  \PH{\trad{\B{n.inp}}{\ListOf{\B{Inp}}}_{\text{decl}}}, 
  \PH{\trad{\B{n.out}}{\ListOf{\B{Out}}}_{\text{decl}}}) \nl"} \\
  \trad{\B{n}}{\B{inp}}_{\text{decl}} \isdef & \trad{\B{n.type}}{\B{Type}} \\
  \trad{\B{n}}{\B{out}}_{\text{decl}} \isdef & \llvm{"\PH{\trad{\B{n.type}}{\B{Type}}}*"}
\end{align*}
It is noteworthy that the outputs of B operations are encoded as pointer
parameters.
\subsection{The implementation section}
